opXRD: Open Experimental Powder X-ray Diffraction Database

### 2024-12-01 chapters/00_abstract.tex
Added dataset overview tables

Powder X-ray diffraction (pXRD) experiments are a cornerstone for materials structure characterization.
Despite their widespread application, the analysis of pXRD diffractograms still presents a significant challenge to automation and thus a bottleneck in high-throughput experimentation of automated materials discovery in self-driving labs.
Machine learning has emerged as a promising research direction to resolve this bottleneck by enabling automated powder diffraction data analysis.
A notable difficulty in applying machine learning to this domain is the lack of sufficiently sized experimental datasets, which has relegated machine learning researchers to train primarily on simulated data. However, models trained on simulated pXRD patterns showed limited generalization ability when applied to experimental patterns, particularly with low-quality experimental pattern that has high noise levels and elevated background.
%Therefore, developing an experimental database offers at least two key advantages: (1) providing a comprehensive benchmark dataset to clarify model performance in practical applications, and (2) advancing phase identification tasks by supplying target-domain data within a transfer learning framework.
With the Open Experimental Powder X-Ray Diffraction Database (opXRD), we aim to provide an openly available and easily accessible dataset of both labeled and unlabeled experimental powder diffractograms, providing machine learning researchers with a large quantity of real experimental diffractograms collected from a broad range of samples. The labeled data can be used as a benchmark to evaluate the performance of models on real experimental data and unlabeled data can help improve the performance of models on experimental data through transfer learning methods. 
%We provide almost barrier-free software tools that allow experimental researchers to find their data and make it accessible - in virtually any widely used format.
We collected XXX labeled and YYY unlabeled diffractograms from a wide spectrum of materials classes, which establishes the first version of the opXRD database.
We hope that this ongoing effort can guide machine learning research toward fully automated analysis of pXRD data and thus enable future self-driving materials labs.

>>>>

Powder X-ray diffraction (pXRD) experiments are a cornerstone for materials structure characterization.
Despite their widespread application, the analysis of pXRD diffractograms still presents a significant challenge to automation and thus a bottleneck in high-throughput materials discovery in self-driving labs.
Machine learning has emerged as a promising research direction to resolve this bottleneck by enabling automated powder diffraction data analysis.
A notable difficulty in applying machine learning to this domain is the lack of sufficiently sized experimental datasets, which has constrained machine learning researchers to train primarily on simulated data. However, models trained on simulated pXRD patterns showed limited generalization ability when applied to experimental patterns, particularly for low-quality experimental patterns with high noise levels and elevated background.
%Therefore, developing an experimental database offers at least two key advantages: (1) providing a comprehensive benchmark dataset to clarify model performance in practical applications, and (2) advancing phase identification tasks by supplying target-domain data within a transfer learning framework.
With the Open Experimental Powder X-Ray Diffraction Database (opXRD), we aim to provide an openly available and easily accessible dataset of both labeled and unlabeled experimental powder diffractograms, providing machine learning researchers with a large quantity of real experimental diffractograms collected from a broad range of materials systems. The labeled data can be used as a benchmark to evaluate the performance of models on real experimental data and the unlabeled data can help improve the performance of models on experimental data, e.g. through transfer learning methods. 
%We provide almost barrier-free software tools that allow experimental researchers to find their data and make it accessible - in virtually any widely used format.
We collected XXX labeled and YYY unlabeled diffractograms from a wide spectrum of materials classes, which establishes the first version of the opXRD database.
We hope that this ongoing effort can guide machine learning research toward fully automated analysis of pXRD data and thus enable future self-driving materials labs.

### 2024-12-01 chapters/01_introduction.tex
Added dataset overview tables

% On Rietveld refinement
After fabricating a new material, a number of analysis techniques can be used to characterize the sample. One method that can be used for phase identification, phase quantification, grain size characterization, and partially to determine the crystal structure of a new material is powder X-ray diffraction. To determine crystal structure from pXRD data, an initial guess is optimized using Rietveld refinement. In this process, an initial crystal structure including unit cell parameters is provided by the operator of the analysis software and fitted to the observed diffractogram in an iterative optimization process based on simulated forward prediction of expected pXRD diffractograms given the current crystal structure\cite{Dinnebier2019, Cano2021}. As Rietveld refinement is a local optimization method, the result of the refinement procedure is only as good as the initial guess it was provided with. \\

% Manual vs (ML) automated Refinement
Manually performing Rietveld refinement is time-consuming and often requires expert knowledge. It is not scalable to the degree required to keep up with advances in throughput and efficiency in other steps of the experimentation pipeline. In a Rietveld refinement process, the operator starts by setting parameters that characterize the background and peak shape of the powder diffraction pattern and determining a crystal structure from which the optimization process can start. The latter is usually done using search-match software, which identifies crystal structures with similar powder diffraction patterns from a database of crystal structures with accompanying powder diffraction patterns. However, attempting to refine all crystal structure parameters at once, starting from this initial structure, is known to lead to unphysical results. What makes the refinement process time-consuming and difficult is finding the correct order in which to refine the crystal structure parameters\cite{Ozaki2020}.  Additionally, if the fit develops into an unphysical set of parameters in this iterative determination of parameters, the operator must intervene to backtrack to the last step. \\

@@

% Predicting structure information with machine learning
Models trained on datasets with simulated diffractograms have already shown very strong performance in predicting lattice parameters\cite{Dong2021, Chitturi2021, Habershon2004, zhang2024crystallographic}, spacegroup \cite{Schopmans2023, Oviedo2018, Park2017, Vecsei2018, Zaloga2020, Suzuki2020, Chakraborty2021,zhang2024crystallographic}, and crystallite size \cite{Dong2021, Chakraborty2021} from simulated diffractograms.
However, the performance substantially drops off when these models are applied to data originating from experiments \cite{Schopmans2023,zhang2024crystallographic}[cite 2-3 papers, incl. ours].
The patterns observed in experiments exhibit imperfections that do not appear when simulating the X-ray diffraction pattern of a powder sample under ideal conditions. The physical effects underlying these imperfections need to be modeled accurately in order to be able to transfer models trained on simulated data to experimental patterns. Most approaches currently model these imperfections as a simple signal based on a heuristic functional form added to the simulated diffractograms.

At the moment, the number of openly available datasets for experimental powder data is very limited. The main sources for experimental patterns are XXX, YYY, and ZZZ. Further commercial datasets such as KKK and LLL exist, but license fees and restrictions in their use to train publishable models or even publish the training data limit their usability. Collecting a labeled dataset of raw experimental powder X-ray diffraction data of sufficient size to directly train a model that predicts crystal structure information is likely out of reach. The most advanced ML models currently are trained on approximately $10^7 - 10^8$ simulated diffractograms, which is - to the best of our knowledge - three orders of magnitude larger than the largest currently curated experimental dataset (PDF-5+ with approximately 20,000 experimental patterns) and even x orders of magnitude larger than the number of unlabeled diffractograms in the opXRD dataset we present here. \\

However, even small labeled and also unlabeled experimental diffractograms hold significant value in practice. First of all, labeled data can be used to better test and benchmark existing and new automated analysis approaches, which yields a better comparison than when using the very limited currently available datasets, such as the commonly used RRUFF mineral dataset\cite{Armbruster2015}. \\

@@

Additionally, some of the observed imperfections, such as background and noise, can be accounted for using transfer learning approaches. In transfer learning the objective is to transfer the capabilities of a model learned on a source domain in which labeled data is abundant to a target domain in which labeled data is sparse\cite{Zhuang2021}. In this context, the source domain is simulated powder diffraction patterns and the target domain is experimental powder diffraction patterns. Many approaches to transfer learning have been proposed, particularly in the domain of image classification \cite{Gatys2016, Ganin2015}. These existing techniques could potentially be adapted to facilitate transfer learning in the context of pXRD patterns.

@@

\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/overview2.png}
    \caption{Overview of the data collection pipeline. Datasets are submitted using an online submission form or our submission helper software. After post-processing and data homogenization, we create a Zenodo entry for each user submission and subsequently include the submission in the opXRD database.}
    \label{fig:overview}
\end{figure*}

% How to contribute? This aspect deserves a separate paragraph
The opXRD database is planned as a growing, community-driven initiative. The dataset we present here is the first version, but we hope to further increase the dataset size through active engagement with the pXRD community. Our primary objective is to minimize the effort and thus the barrier to contributing experimental data to the opXRD database. Thus, we developed a program that helps to find and share data from e.g. pXRD lab computers. In detail, users can select their most common pXRD file types, the program lists all files of that type, and users can select or deselect certain folders or files for sharing. Selected contributions will be uploaded to opXRD, processed to a common file format, and - if wanted - published on Zenodo on behalf of the contributors, before becoming part of the opXRD database. If labels are available, they can be shared with opXRD as well. Further details can be found on the opXRD website (Figure~\ref{fig:overview}). \\

As argued by Aranda and Kroon-Batenburg \textit{et al}.\cite{Aranda2018, Kroon-Batenburg2024}, sharing raw powder diffraction data is not only in the interest of furthering machine learning research but is also in line with open science principles. It furthers the ability of other researchers to reproduce published work and in turn, adds to the credibility of the publisher of the data. Considering our streamlined submission and post-processing framework, a publication of powder diffraction data through the opXRD database is a simple way of realizing these principles. Compared to uploading the data to Zenodo it also has the added benefit that the data is contributed towards a large, homogenous dataset with a standardized interface. This makes the data more easily accessible to other researchers and provides more value to researchers seeking great quantities of data compared to an isolated submission on Zenodo. \\

The broad range of available experimental samples contained in the opXRD v1.0 database makes it possible to apply state-of-the-art ML approaches to the domain of powder XRD analysis. We hope that this can drive ML research in this field towards more advanced automated analysis workflows that can accelerate material science research through ready application in high-throughput experimentation pipelines.

>>>>

% On Rietveld refinement
After fabricating a new material, a number of analysis techniques can be used to characterize the sample. One method that can be used for phase identification, phase quantification, grain size characterization, and to determine the crystal structure of a new material is powder X-ray diffraction. To determine the crystal structure from pXRD data, an initial guess is typically optimized using Rietveld refinement. In this process, an initial crystal structure including unit cell parameters is provided by the researcher and subsequently fitted to the observed diffractogram in an iterative optimization process based on simulated forward prediction of pXRD diffractograms given the current crystal structure\cite{Dinnebier2019, Cano2021}. As Rietveld refinement is a local optimization method, the result of the refinement procedure is only as good as the initial guess it was provided with. \\

% Manual vs (ML) automated Refinement
Manually performing Rietveld refinement is time-consuming and often requires expert knowledge. It is not scalable to the degree required to keep up with advances in throughput and efficiency in other steps of the experimentation pipeline. In a Rietveld refinement process, the operator starts by setting parameters that characterize the background and peak shape of the powder diffraction pattern and then determines a crystal structure from which the optimization process can start. The latter is usually done using search-match software, which identifies crystal structures with similar powder diffraction patterns from a database of crystal structures with accompanying powder diffraction patterns. However, attempting to refine all crystal structure parameters at once, starting from this initial structure, is known to lead to unphysical results. What makes the refinement process time-consuming and difficult is finding the correct order in which to refine the crystal structure parameters\cite{Ozaki2020}. \\ %Additionally, if the fit develops into an unphysical set of parameters in this iterative determination of parameters, the operator must intervene to backtrack to the last step. \\

@@

% Predicting structure information with machine learning
Models trained on datasets with simulated diffractograms have already shown very strong performance in predicting lattice parameters\cite{Dong2021, Chitturi2021, Habershon2004, zhang2024crystallographic}, spacegroup \cite{Schopmans2023, Oviedo2018, Park2017, Vecsei2018, Zaloga2020, Suzuki2020, Chakraborty2021,zhang2024crystallographic}, and crystallite size \cite{Dong2021, Chakraborty2021} from simulated diffractograms.
However, the performance substantially drops off when these models are applied to data originating from experiments \cite{Schopmans2023,zhang2024crystallographic}[cite 2-3 papers, incl. ours].
The patterns observed in experiments exhibit imperfections that do not appear when simulating the X-ray diffraction pattern of a powder sample under ideal conditions. The physical effects underlying these imperfections need to be modeled accurately in order to be able to transfer models trained on simulated data to experimental patterns. Most approaches currently model these imperfections as a simple signal based on a heuristic functional form added to the simulated diffractograms.\\

At the moment, the number of openly available datasets for experimental powder data is very limited. The main sources for experimental patterns are XXX, YYY, and ZZZ. Further commercial datasets such as KKK and LLL exist, but license fees and restrictions in their use to train publishable models or even publish the training data limit their usability. Collecting a labeled dataset of raw experimental powder X-ray diffraction data of sufficient size to directly train a model that predicts crystal structure information is currently out of reach. The most advanced ML models currently are trained on approximately $10^7 - 10^8$ simulated diffractograms, which is - to the best of our knowledge - three orders of magnitude larger than the largest currently curated experimental dataset (PDF-5+ with approximately 20,000 experimental patterns) and even x orders of magnitude larger than the number of unlabeled diffractograms in the opXRD dataset we present here. \\

However, even small labeled and also unlabeled experimental diffractograms hold significant value in practice. First of all, labeled data can be used to test and benchmark existing and new automated analysis approaches, which yields a better comparison than when using the very limited currently available datasets, such as the commonly used RRUFF mineral dataset\cite{Armbruster2015}. \\

@@

Additionally, some of the observed imperfections can be accounted for using transfer learning approaches. In transfer learning the objective is to transfer the capabilities of a model learned on a source domain in which labeled data is abundant to a target domain in which labeled data is sparse\cite{Zhuang2021}. In this context, the source domain is simulated powder diffraction patterns and the target domain is experimental powder diffraction patterns. Many approaches to transfer learning have been proposed, particularly in the domain of image classification \cite{Gatys2016, Ganin2015}. These existing techniques can be adapted to facilitate transfer learning in the context of pXRD patterns.\\

@@

\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\linewidth]{figures/overview.png}
    \caption{Overview of the data collection pipeline. Datasets are submitted using an online submission form, optionally with the help of our submission helper software. After post-processing and data homogenization, we offer creation of a Zenodo entry for each user submission and subsequently include the submission in the opXRD database.}
    \label{fig:overview}
\end{figure*}

% How to contribute? This aspect deserves a separate paragraph
The opXRD database is intended as a growing, community-driven initiative. The dataset we present here is the first version, but we hope to further increase the dataset size through active engagement with the pXRD community. Our primary objective is to minimize the effort and thus the barrier to contributing experimental data to the opXRD database. Thus, we developed a program that helps to find and share data from pXRD lab computers. In detail, users can select their most common pXRD file types, the program lists all files of that type, and users can select or deselect certain folders or files for sharing. Selected contributions will be uploaded to opXRD, processed to a common file format, and - if wanted - published on Zenodo on behalf of the contributors, before becoming part of the opXRD database. If labels are available, they can be shared with opXRD as well. Further details can be found on the opXRD website (Figure~\ref{fig:overview}). \\

As argued by Aranda and Kroon-Batenburg \textit{et al}.\cite{Aranda2018, Kroon-Batenburg2024}, sharing raw powder diffraction data is not only in the interest of furthering machine learning research but is also in line with open science principles. It furthers the ability of other researchers to reproduce published work and in turn, adds to the credibility of the publisher of the data. Considering our streamlined submission and post-processing framework, a publication of powder diffraction data through the opXRD database is a simple way of realizing these principles. Compared to publishing data individually, it also has the added benefit that the data is contributed towards a large, homogenous dataset with a standardized interface. This makes the data more easily accessible to other researchers and provides more value to researchers seeking large quantities of data compared to an isolated submission on Zenodo. However, further data annotation with metadata is required to fulfill the FAIR data principles.\\

The broad range of available experimental samples contained in the opXRD v1.0 database makes it possible to apply state-of-the-art ML approaches to the domain of powder XRD analysis. We hope that this can drive ML research in this field towards more advanced automated analysis workflows that can accelerate materials science research through ready application in high-throughput experimentation pipelines.

### 2024-12-01 chapters/03_SOTA_ML.tex
Added dataset overview tables

Using simulated diffractograms based on structures from the ICSD, Lee {\it et al.} trained a deep convolutional neural network (CNN) that is able to classify occurring phases in diffractograms of a specific compound pool \cite{Lee2020}. They furthermore developed models based on fully convolutional neural networks and transformer encoders that predict the crystal system and the spacegroup and other structural properties, such as the band gap \cite{Lee2022}. With their best model for the crystal system prediction on ICSD structures, they achieved a test accuracy of \SI{92.2}{\percent}. Park {\it et al.} reached a test accuracy of roughly 81\% for a CNN, which classifies space groups of simulated single-phase diffractograms \cite{Park2017}. \\

A regression analysis on lattice parameters within a broader framework encompassing all material classes was conducted by Chitturi et al\cite{Chitturi2021}. They developed a distinct CNN for each crystal system, utilizing a merged dataset from both the ICSD and the Cambridge Structural Database, and managed to achieve a mean absolute percentage error of about \SI{10}{\percent} for the lattice lengths, although they encountered difficulties in accurately predicting angles.
Zhang {\it et al.} introduce a convolutional self-attention neural network trained on simulated patterns to classify crystal types \cite{zhang2024crystallographic}. Their model was tested on 23,073 unary, binary, and ternary inorganic crystal structures sourced from the COD. The study observed a noticeable performance drop when the pre-trained model was applied to real experimental patterns, as opposed to simulated data. However, their recent work \cite{cao2024simxrd} proposes using convolutional peak descriptors that consider the detector's geometry, which significantly reduces the performance gap in their benchmark tests.\\

@@

TODO: We need a short literature review here; needs to also cover papers published after our previous paper!

@@

TODO: briefly discuss papers about multiphase classification and remixing of diffraction patterns
Non-negative Matrix Factorization 2018: \url{https://link.springer.com/chapter/10.1007/978-3-319-93031-2_4}
non-negative matrix factorization 2018: \url{https://ojs.aaai.org/aimagazine/index.php/aimagazine/article/view/2785}
Physically-informed Graph-based DRNet (PG-DRNet) 2023: \url{https://ieeexplore.ieee.org/abstract/document/10459774/}

TODO: discuss in more detail why domain randomization as we did in the last paper so far failed for background and noise, and thus why we need background and noise statistics from experiment to enhance simulated datasets and transfer them to the experimental data distribution.

>>>>

Using simulated diffractograms based on structures from the ICSD, Lee {\it et al.} trained a deep convolutional neural network (CNN) that is able to classify occurring phases in diffractograms of a specific compound pool \cite{Lee2020}. They furthermore developed models based on fully convolutional neural networks and transformer encoders that predict the crystal system and the spacegroup and other structural properties, such as the band gap \cite{Lee2022}. With their best model for the crystal system prediction on ICSD structures, they achieved a test accuracy of \SI{92.2}{\percent}. Park {\it et al.} reached a test accuracy of roughly \SI{81}{\percent} for a CNN, which classifies space groups of simulated single-phase diffractograms \cite{Park2017}. \\

A regression analysis on lattice parameters within a broader framework encompassing all material classes was conducted by Chitturi {\it et al} \cite{Chitturi2021}. They developed a distinct CNN for each crystal system, utilizing a merged dataset from both the ICSD and the Cambridge Structural Database, and managed to achieve a mean absolute percentage error of about \SI{10}{\percent} for the lattice lengths, although they encountered difficulties in accurately predicting angles.
Zhang {\it et al.} introduce a convolutional self-attention neural network trained on simulated patterns to classify crystal types \cite{zhang2024crystallographic}. Their model was tested on 23,073 unary, binary, and ternary inorganic crystal structures sourced from the COD. The study observed a noticeable performance drop when the pre-trained model was applied to real experimental patterns, as opposed to simulated data. However, their recent work \cite{cao2024simxrd} proposes using convolutional peak descriptors that consider the detector's geometry, which significantly reduces the performance gap in their benchmark tests.\\

@@

Schuetzke {\it et al.} trained a classifier to classify if a diffractogram stems from an amorphous, single-phase, or multi-phase sample \cite{Schuetzke2024}. Due to the lack of experimental pXRDs, they build a pipeline to augment simulated diffractograms by, among other things, slightly varying the underlying crystal lattice.

In our 2023 paper, we came up with the approach to generate crystal structures and their corresponding pXRD patterns on the fly during the training process \cite{Schopmans2023}. This approach defeats the issue of
a limited dataset size, which limits the depth of neural networks that can be trained. However, when we applied our space group classification model to experimental patterns from the RRUFF database the accuracy dropped substantially. Augmenting our simulated patterns with background, noise, and impurities would help to bring simulated diffractograms closer to experimental ones, making models trained on them more performant on experimental diffractograms. For this augmentation, background and noise statistics from a broader experimental pXRD database would be very valuable.

@@

Physically-informed Graph-based DRNet (PG-DRNet) 2023: \url{https://ieeexplore.ieee.org/abstract/document/10459774/}

### 2024-12-01 chapters/04_our_dataset.tex
Added dataset overview tables

@@

%% Paragraph by FX Coudert and Arthur Hardiagon
\subsubsection*{Chimie Paris Tech, PSL}

>>>>

\begin{table}[!htb]
\centering
\caption{Labeled opXRD database entries}
\label{tab:labeled}
\begin{tabular}{@{}ccc@{}} % Adjusted columns
\toprule
Institution & Research project & No. patterns\footnote{This number is subject to change} \\
\midrule
CNRS        & Extracted from the Crystallography open database & 895 \\
USC         & Study of CuNi and CuAl alloys & 338 \\
HKUST(GZ)   & Reference database for phase identification & 21 \\
EMPA        & Tin halide perovskites and combinatorial Zn-V-N library & 790 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[!htb]
\centering
\caption{Unlabeled opXRD database entries}
\label{tab:unlabeled}
\begin{tabular}{@{}ccc@{}} % Adjusted columns
\toprule
Institution & Research project & No. patterns\footnote{This number is subject to change} \\
\midrule
INT         & Compilation of various research projects & 19780 \\
IKFT        & Compilation of various research projects & 64 \\
LBNL        & Time series recording of in-situ measurements & n/a \\
\bottomrule
\end{tabular}
\end{table}

@@

%% Paragraph by FX Coudert and Arthur Hardiagon
\subsubsection*{Chimie ParisTech, PSL University}

### 2024-12-01 chapters/05_how_to_use.tex
Added dataset overview tables

TODO: A small tutorial with a few python commands to show how easy it is to use our dataset through the provided API.

A major advantage of this dataset over other comparable datasets is that it is very easy to handle in python and in PyTorch.
Our accompanying python library xrdpattern (\url{https://github.com/aimat-lab/xrdpattern}) provides a means to read the dataset simply by using PatternDB.load(data\_dirpath). The patterns attribute of this class is a list of the indivdual pattern diffractograms, each of which supports a standardize, plot and to tensor method. \\
The standardize method returns a "standardized" version of the pattern with a fixed angle range, a fixed number of entries and intensities normalized to the $[0,1]$ interval.

% caobin
We have launched Xqueryer (http://xqueryer.caobin.asia/), a free online platform for phase identification. Users can upload experimental PXRD data to identify and match structures from the Materials Project database. Uploaded data is stored and integrated into our opXRD database.

>>>>

Next to the availability of the opXRD dataset on Zenodo, we further provide an easy-to-use Python library (\url{https://github.com/aimat-lab/xrdpattern}) to download and interface the dataset.
The dataset can be accessed using the simple command 
\pyth{PatternDB.load(data_dirpath)}. Our tool further includes options for standardization, plotting, and the conversion to \emph{PyTorch} tensors.
We provide the following Jupyter notebook that showcases the functionality of the library in more detail: <Link to github Jupyter notebook on Colab>

%The patterns attribute of this class is a list of the indivdual pattern diffractograms, each of which supports a standardize, plot and to tensor method. The standardize method returns a "standardized" version of the pattern with a fixed angle range, a fixed number of entries and intensities normalized to the $[0,1]$ interval.
